---
title: "04_risk_scoring_typology"
author: "Tamara"
date: "2025-06-14"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(here)
knitr::opts_chunk$set(echo = FALSE)
```


## Loading the data (merged)

```{r}

df_clean <- read_excel(here("data_clean", "merged_attendance_eqi.xlsx"))

glimpse(df_clean)
```

# STEP 1: Standardise predictors
# Standardise EQI and Volatility to z-scores
```{r}
df_clean <- df_clean %>%
  mutate(
    eqi_std = as.numeric(scale(eqi_mean)),
    vol_std = as.numeric(scale(volatility_present))
  )
```
# STEP 2: Assigning weighted risk score based on variance explained in 03_modeling_risk_factors (~63% EQI, ~37% volatility)
```{r}
df_clean <- df_clean %>%
  mutate(
    risk_score = (0.63 * eqi_std) + (0.37 * vol_std)
  )
```
# STEP 3: Quick check of risk score distribution
```{r}
summary(df_clean$risk_score)
```

# STEP 4: Validating the association between risk score and attendance using a linear model

```{r}
model_lin <- lm(avg_present ~ risk_score, data = df_clean)
summary(model_lin)
```

# STEP 5: Visualise relationship

```{r, warning=FALSE}
ggplot(df_clean, aes(x = risk_score, y = avg_present)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Attendance vs Weighted Risk Score",
       x = "Weighted Risk Score",
       y = "Average Attendance") +
  theme_minimal()
```

# STEP 6: Building the typology classification. 
## Thresholds were chosen as ±0.5 standard deviations from the mean, capturing regions meaningfully above or below average. This results in four interpretable categories:

### High risk (structural + instability): High EQI disadvantage and high attendance volatility.
### Structural risk: Disadvantaged but attendance is relatively stable.
### Instability risk: Less disadvantaged, but engagement fluctuates.
### Lower risk: Relatively advantaged and stable attendance.

```{r}
df_clean <- df_clean %>%
  mutate(
    risk_typology = case_when(
      eqi_std <= -0.5 & vol_std >= 0.5  ~ "High Risk (Structural + Instability)",
      eqi_std <= -0.5 & vol_std < 0.5   ~ "Structural Risk",
      eqi_std > -0.5 & vol_std >= 0.5   ~ "Instability Risk",
      eqi_std > -0.5 & vol_std < 0.5    ~ "Lower Risk"
    ),
    risk_typology = factor(risk_typology, levels = c("High Risk (Structural + Instability)", "Structural Risk", "Instability Risk", "Lower Risk"))
  )
```

# STEP 7: Summary Table for README (NA values are Auckland and All)
```{r}
typology_summary <- df_clean %>%
  group_by(risk_typology) %>%
  summarise(
    mean_attendance = round(mean(avg_present), 1),
    count = n()
  )
typology_summary
```

# STEP 8: Visualization of the typology from step 6
```{r, warning=FALSE}
ggplot(df_clean, aes(x = eqi_mean, y = volatility_present, color = risk_typology, label = education_region)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_text(size = 3, vjust = -1, check_overlap = TRUE) +
  labs(
    title = "Regional Risk Typologies",
    x = "EQI (Mean)",
    y = "Volatility",
    color = "Risk Typology"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set2")
```